Estimating the number of clusters using Cross Validation

### 2015-06-23 cvclust.tex
fix paragraph spacing in introduction

\section{Introduction}
As a main task of exploratory data analysis, clustering organizes unlabeled observations into groups such that observations in same group are more similar compare to those in different group. Clustering is an important topic in unsupervised learning because it can reveal the internal structure of data through grouping, segment the data through partitioning and summarize data for other purposes such as dimension reduction. It has being widely used in various fields such as psychology, biology, statistics and machine learning including pattern recognition, image segmentation etc.\\

@@

Recently, there are several new proposals to find the $k$ automatically. Gap statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the change in within-cluster dispersion with that expected under an appropriate reference null distribution. Specifically, the graph of $log(W_k)$ is compared with its expectation under an appropriate null reference distribution of the data. The value of $k$ associated with the largest gap between $log(W_k)$ and the reference curve is selected as optimal $k$. \cite{sugar2003finding} proposed an approach which finds the number of clusters based on distortion, a quantity that measures the average distance, per dimension. It's backed by a rigorous theoretical justification based on information-theoretic ideas.   \cite{fraley2002model}'s Model-based method employs the EM algorithm to estimate the parameters in Gaussian mixture model, and select the best model ($k$) using BIC criterion. Stability-based criterion is also proposed to locate the best $k$ by some authors such as \cite{ben2001stability}, \cite{wang2010consistent} and \cite{fang2012selection}. \cite{chiang2010intelligent} provides a nice review of existing methods for finding the right $k$ in published literature. \\

Most existing methods are either model based method requires strong parameter assumptions or ad hoc method utilizes within-cluster dispersion. Although many view selecting the number of clusters as a model selection problem, very few approaches this problem from the prediction point of view. Select model with smallest prediction error via cross-validation is one of the simplest and most widely used model selection techniques in supervised learning. The lack of true class (label) in data set makes the adoption of cross-validation into unsupervised leaning problem difficult. Perhaps the only exception is \cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction strength. The strategy is to first cluster the test data and training data into $k$ clusters respectively. Then, for each pair of observations that are assigned to the same test cluster, algorithm determines whether they are also assigned to the same cluster based on the training centers. The intuition here is, if $k=k_0$, the true number of clusters, then the $k$ training set clusters will be similar to the $k$ test clusters, and hence will predict them well. However, a specifically defined prediction error measure is used in such procedure, which is quite different from the one commonly used in cross-validation procedure in supervised learning. Therefore the well understood properties of cross-validation procedure in supervised learning cannot be carried over to the prediction strength method, makes the interpretation of its result difficult. \\

>>>>

\section{Introduction}
As a main task of exploratory data analysis, clustering organizes unlabeled observations into groups such that observations in same group are more similar compare to those in different group. Clustering is an important topic in unsupervised learning because it can reveal the internal structure of data through grouping, segment the data through partitioning and summarize data for other purposes such as dimension reduction. It has being widely used in various fields such as psychology, biology, statistics and machine learning including pattern recognition, image segmentation etc.

@@

Recently, there are several new proposals to find the $k$ automatically. Gap statistics \citep{tibshirani2001estimating} estimates $k$ by comparing the change in within-cluster dispersion with that expected under an appropriate reference null distribution. Specifically, the graph of $log(W_k)$ is compared with its expectation under an appropriate null reference distribution of the data. The value of $k$ associated with the largest gap between $log(W_k)$ and the reference curve is selected as optimal $k$. \cite{sugar2003finding} proposed an approach which finds the number of clusters based on distortion, a quantity that measures the average distance, per dimension. It's backed by a rigorous theoretical justification based on information-theoretic ideas.   \cite{fraley2002model}'s Model-based method employs the EM algorithm to estimate the parameters in Gaussian mixture model, and select the best model ($k$) using BIC criterion. Stability-based criterion is also proposed to locate the best $k$ by some authors such as \cite{ben2001stability}, \cite{wang2010consistent} and \cite{fang2012selection}. \cite{chiang2010intelligent} provides a nice review of existing methods for finding the right $k$ in published literature.

Most existing methods are either model based method requires strong parameter assumptions or ad hoc method utilizes within-cluster dispersion. Although many view selecting the number of clusters as a model selection problem, very few approaches this problem from the prediction point of view. Select model with smallest prediction error via cross-validation is one of the simplest and most widely used model selection techniques in supervised learning. The lack of true class (label) in data set makes the adoption of cross-validation into unsupervised leaning problem difficult. Perhaps the only exception is \cite{tibshirani2005cluster}, which selects the optimal $k$ by prediction strength. The strategy is to first cluster the test data and training data into $k$ clusters respectively. Then, for each pair of observations that are assigned to the same test cluster, algorithm determines whether they are also assigned to the same cluster based on the training centers. The intuition here is, if $k=k_0$, the true number of clusters, then the $k$ training set clusters will be similar to the $k$ test clusters, and hence will predict them well. However, a specifically defined prediction error measure is used in such procedure, which is quite different from the one commonly used in cross-validation procedure in supervised learning. Therefore the well understood properties of cross-validation procedure in supervised learning cannot be carried over to the prediction strength method, makes the interpretation of its result difficult.
